---
title: "The Proof of MAIC"
author:
  - name: Nam-Anh
date: "`r Sys.Date()`"
bibliography: citation.bib
csl: citation.csl
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    pandoc_args: ["--number-sections"]
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

\newcommand{\tp}[1]{{#1}^{\top}}
\newcommand{\bf}[1]{\boldsymbol{#1}}


```{r setup, include=FALSE}
if(!library(pacman, logical.return = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, rjags, magrittr, kableExtra)

MyTable = function(x, cap = NULL, scroll_box = FALSE, ...){
      result<-
          kableExtra::kbl(x, booktab = TRUE, caption = cap,...)|>
          kableExtra::kable_styling(bootstrap_options = c("striped"))
     
     if(scroll_box){
          
          css = c("border:0.1px;", 
                  "margin-left: auto;", 
                  "margin-right: auto;", 
                  "margin-bottom: 25px;")
          result|>  
               kableExtra::scroll_box(height = "500px",
                                      extra_css  = paste0(css, collapse = ""))
     } else result
}
```

```{css, include = FALSE}
.makebox{
border: 2px solid #566573;
border-radius: 5px; 
margin-bottom: 15px;
}
```


Let $\bf{p} = \tp{(p_1,p_2,\dots,p_n)}$ be propensity score and $\bf X = \tp{(\bf{x}_1,\bf{x}_2, \dots, \bf{x}_n)}$ baseline characteristics of subjects in IPD. Also, $\bf{\mu}$ denotes average vector of baseline characteristics found in SLD. We aim to estimate $\bf p$ such that $\tp{\bf{p}}\bf{X} = \tp{\bf{\mu}}$, i.e. we want to minimize *Kullback Leibler divergence* of $\bf p$ and empirical function with some constraints. Specifically, we want to minimize 

$$
\mathcal{L} = -\tp{\bf{p}}\ln(\frac{1}{n\bf{p}}),
$$

with two constraints:

$$
\tp{\bf{p}}\bf{1} = 1, \quad and \quad \tp{\bf{p}}\bf{X} = \tp{\bf{\mu}}.
$$ 

Alternatively, we optimize 

$$
\mathcal{L} = -\tp{\bf{p}}\ln(\frac{1}{n\bf{p}}) - \alpha_0\Big(\sum_{i=1}^np_i -1\Big) - \alpha_1(p^{\top}X-\boldsymbol{\mu}^{\top})
(\#eq:eq1)
$$

Equation \@ref(eq:eq1) could be solved by taking derivative w.r.t $\bf{p}$, $\alpha_0$ and $\alpha_1$ followed by equating derivations to zero and solve equations. 

Taking derivative w.r.t $p_i$:

$$
\begin{aligned}
&\quad \frac{\partial\mathcal{L}}{\partial p_i} = 0 \\
&\Rightarrow \ln(np_i) +n -\alpha_0 -\alpha_1(x_i-\mu) = 0 \\
&\Rightarrow \ln(np_i) = \alpha_1(x_i-\mu)+\alpha_0 -n \\
&\Rightarrow p_i = \frac{\exp[\alpha_0+\alpha_1(x_i -\mu)-n]}{n} \\
&\Rightarrow p_i = \frac{e^{-n}}{n}\exp[\alpha_0+ \alpha_1(x_i-\mu)]
\end{aligned}
(\#eq:eq2)
$$

Also,

$$
\begin{aligned}
&\quad 1 = \sum_{i=1}^np_i = \frac{e^{-n}}{n}\sum_{i=1}^n\exp[\alpha_0+\alpha_1(x_i-\mu)] \\
& \Rightarrow \frac{n}{e^{-n}} = \sum_{i=1}^n\exp[\alpha_0+ \alpha_1(x_i-\mu)]
\end{aligned}
(\#eq:eq3)
$$

substitute \@ref(eq:eq3) to \@ref(eq:eq2), we obtain 

::: {.makebox}
\begin{equation}
p_i = \frac{\exp[\alpha_0+\alpha_1(x_i - \mu)]}{\sum_{i=1}^n\exp[\alpha_0 +\alpha_1(x_i-\mu)]}
(\#eq:eq4)
\end{equation}
:::

Let $z_i = x_i-\mu$, and hence $p_i = \frac{\exp[\alpha_1 z_i]}{\sum_{i=1}^n\exp[\alpha_1 z_i]}$, implying the weight

$$
w_i = p_i\sum_{j=1}^n\exp[\alpha_1z_j] = \exp (\alpha_1z_i)> 0.
$$

Thus,

$$
\sum_{i=1}^nw_iz_i = \sum_{i=1}^n\Big(p_i\sum_{j=1}^n\exp[\alpha_1z_j]\Big)z_i = \sum_{i=1}^n\exp (\alpha_1z_i)> 0.
(\#eq:eq5)
$$

Recall that $z_i = x_i-\mu$ and we are interested in $\sum_{i=1}^np_ix_i = \mu$ implying that 

$$
\sum_{i=1}^n p_iz_i= 0.
(\#eq:eq6)
$$

Also, the second term in \@ref(eq:eq5) can be rewritten as 

$$
k\sum_{i=1}^np_iz_i
(\#eq:eq7)
$$

where $k = \sum_{j=1}^n\exp[\alpha_1z_j]$. Thus, the condition \@ref(eq:eq6) is satisfied as the equation \@ref(eq:eq7) equals zero, which is equivalent to minimizing the right hand side of \@ref(eq:eq5), $\sum_{i=1}^n\exp (\alpha_1z_i)\ge 0$, which can be rewritten in the matrix form when there are more than one effect modifiers 

$$
\boldsymbol{1}^{\top}\exp(\boldsymbol{Z\alpha}_1) = \sum_{i=1}^n\exp(\boldsymbol{z}_i^{\top}\boldsymbol{\alpha}_1).
$$

For the sake of illustration, we shall consider the following example. Suppose we have effect modifiers age and gender of 100 patients as follows

```{r}
set.seed(1234)
age = rnorm(100,50,5)
gender = sample(c(0,1),100, replace = T)

data.frame(patient= 1:100,age = age, gender = gender)|>
     MyTable(scroll_box = T, cap = "Age and Gender of 100 patients. For *Gender*: 1: Female; 0: Male.")
```

Mean of `age` and proportion of `female` are $\hat{\theta}_{age}$ = `r mean(age)` and $\hat{\theta}_{sex}$ = `r mean(gender)`, respectively. We shall match both $\hat{\theta}_{age}$ and $\hat{\theta}_{sex}$ to `r age2=52;age2` and `r gender2=0.5;gender2`. To this end, we follow above instruction to determine $\alpha_1$ and $p_i$. Thus,

```{r}
Z = matrix(c(age-age2,gender-gender2), ncol = 2)
objfn = function(alpha,Z) sum(exp(Z%*%alpha))
alpha = optim(par = c(0,0), fn = objfn, Z = Z)$par
p = exp(Z%*%alpha)|>
     {\(x) x/sum(x)}()

data.frame(patient= 1:100,age = age, gender = gender, p = p)|>
     MyTable(scroll_box = T, cap = "Age and Gender and MAIC proportion of 100 patients.")
```

Using `p` to calculate weighted mean and proportion of `age` and `gender`,

```{r}
t(p)%*%matrix(c(age,gender),ncol=2)
```

We also calculate effective sample size

```{r}
w = c(exp(Z%*%alpha))
ess = sum(w)^2/sum(w^2)
ess
```

Thus, to balance effect modifiers, we lose about `r ceiling(100 - ess)` participants.






