---
title: "Statistical cost-effectiveness"
author:
  - name: Nam-Anh
date: "`r Sys.Date()`"
bibliography: citation.bib
csl: citation.csl
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    pandoc_args: ["--number-sections"]
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

\newcommand{\bf}[1]{\boldsymbol{#1}}


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

if(!library(pacman, logical.return = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, rjags, magrittr, xaringan, kableExtra)

lb = function(...){
     x = unlist(strsplit(as.character(rlang::expr(...)),"\\."))
     word = switch(which(c("tab","fig","eq")%in% x[1]),"Table ","Figure ","Eq.")
     paste0(word,"\\@ref(",x[1],":",x[2],")")
}

MyTable = function(x, cap = NULL, ...){
     kableExtra::kbl(x, booktab = T, caption = cap,...)%>%
          kable_styling(latex_options = "striped")
}
```

```{css, echo=FALSE}
.makebox{
border: 2px solid #566573;
border-radius: 5px; 
margin-bottom: 15px;
}
```

# Mathematical background

Let $t \in \mathcal{T}$ be a decision or intervention/treatment, and $o \in \mathcal{O}$ consequence or outcome. We also introduce the random quantity $\boldsymbol{\omega} \in \Omega$, so the outcome can be described by relationship $o = (\boldsymbol{\omega},t)$ which could be interpreted as the result of selecting treatment $t$ with a series of random quantities obtained in the future. 

Since $\boldsymbol{\omega}$ is random component, we have the probability measure $p(\boldsymbol{\omega})$, i.e. $p: \Omega \rightarrow [0,1]$. Also, the value of each outcome is quantified by measure of utility $u: \mathcal{O} \rightarrow \mathbb{R}$. In general, the random component is nothing but distribution, i.e. $\boldsymbol{\omega} = (y,\boldsymbol{\theta})$ where $y \in \mathcal{Y}$ is future result of individual(s), and $\boldsymbol{\theta} \in \mathbb{\Theta}$ is a set of parameters. Having had sub-components of $\boldsymbol{\omega}$, the outcome can be rewritten as 

$$
\begin{aligned}
&o = (y,\boldsymbol{\theta},t), \text{ implying that } \\
&p(o) = u(y,\boldsymbol{\theta},t)
\end{aligned}
$$
where 

$$
p(y,\boldsymbol{\theta}) = p(y|\boldsymbol{\theta})p(\boldsymbol{\theta}).
$$

For example, we shall consider 5 states of happiness denoted by $\{y_i\}_{i=1}^5$, each state has probability $\boldsymbol{\theta}_i^t, t = 0,1$. We temporarily assume that the degree of uncertainty of $\boldsymbol{\theta}^t_i$ equal zero. Thus, The preffered outcome is can be obtained in either of two ways

1. directly, base on the probability of first state of each treatment, i.e. $\boldsymbol{\theta}^t_1$.
2. indirectly, through other states. In fact, state $\{y_{s}\}_{s=2}^5$ is often defined as the utilities equivalent to a scenario in which perfect health is obtained with probability $\pi_s = u(y_s,t)$.

Using the law of total probability we obtain the chance of preferred consequence $y_1$ as follows

$$
\begin{aligned}
\mathbb{P}(y_1|\boldsymbol{\theta}^t) &= \boldsymbol{\theta}_1^t + \sum_{s=2}^5\pi_s\mathbb{P}(y_s|\boldsymbol{\theta}^t)
\\&= \boldsymbol{\theta}_1^t+ \sum_{s=1}^5\pi_s\boldsymbol{\theta}_s^t
\\&=\sum_{s=1}^5u(y_s,t)\boldsymbol{\theta}_s^t
\\&= \mathbb{E}[u(Y,t)]
\end{aligned}
(\#eq:1)
$$

In practice, $\boldsymbol{\theta}^t$ always involves uncertainty, so we will need to consider $p(\boldsymbol{\theta}^t|\mathcal{D})$, where $\mathcal{D}$ is the background information. The implication of `r lb(eq.1)` is that the probability of obtaining the preferred options amounts to the expectation of utility function, more commonly referred to as the **expected utility** denoting $\mathcal{U}^t$. Thus, the optimal decision criterion is to maximize the expected utility, i.e.

:::{.makebox}
$$
t^* = \arg\max_{\mathcal{T}}\mathcal{U}^t
$$
:::

Also, we can generalize `r lb(eq.1)` as follows

$$
\mathcal{U}^t = \int_{\Theta}\int_{\mathcal{Y}} u(y,t)p(y|\boldsymbol{\theta}^t)p(\boldsymbol{\theta}^t|\mathcal{D})dyd\boldsymbol{\theta}^t
(\#eq:2)
$$

# The setup of decision-making in health economics

## Statistical framework

Let $t \in \mathcal{T}:= \{0,1\}$ be a set of interventions and $y_i$ outcome of $i^{th}$ subject which is constituted by effectiveness and cost, i.e. $y = (e,c)$. Also, $\bf{\theta} = (\theta_0,\theta_1)^{\top}$ is the vector of parameters of interest. First, we aim to find the conditional distribution of $\bf\theta$ given data, i.e. posterior distribution of $\bf\theta$. Let $\mathcal{D}_t = \{y_i\}_{i=1}^{n_t}$ and $\mathcal{D} = \bigcup_{t\in\mathcal{T}}\mathcal{D}_t$, so the posterior has the form $p(\bf{\theta}|\mathcal{D})$. In the second step, therefore, we are interested in predictive model which can be obtained by simulation.

For example, suppose that $\pi_1$ and $\pi_0$ are probabilities that a patient is treated by active treatment and placebo, respectively, and $\pi_1 = \pi_0\rho$ where $\rho$ is the measure of difference, i.e absolute or relative difference. Conditionally on $\pi_t$, $\gamma$ is the probability of needing ambulatory care and $1-\gamma$ is the probability a patient requires hospital admission. 

We take the sample size of $N =1000$ and the number of patients experiencing side effect is $SE_t \sim \mathcal{Bin}(\pi_t,N)$; the number of patients requiring ambulatory care is $A_t \sim \mathcal{Bin}(\gamma,SE_t)$, and hence the number of patients with side effects requring hospital admission is $H_t = SE_t-A_t$. Also, $c_t^{drug}$ is a fixed cost value and $c^{hosp}$ and $c^{amb}$ are costs of ambulatory care and hospital admission, such two costs are independent. We now modify the parameters of interest as $\bf{\theta} = (\bf{\theta}_0,\bf{\theta}_1)$ where $\bf{\theta}_0 = (\pi_0,\gamma,c^{amb},c^{hosp})$ and $\bf{\theta}_1 = (\pi_1,\gamma,c^{amb},c^{hosp})$. Therefore, condition on $\bf\theta$ the observable outcome is $Y = (SE_t,A_t,H_t)$. 

## Decision process

Suppose an intervention $t$ is applied and we obtain the outcome $y$. Thus, we can quantify by combining a clinical effectiveness $e$ of the outcome $y$ and the costs $c$ associated with intervention $t$, for each $(y,t)$ we associate $(e,c)$. We now let 

$$
\Delta_e := \mathbb{E}(e|\theta_1) - \mathbb{E}(e|\theta_0), \text{ and}\quad \Delta_c:=\mathbb{E}(c|\theta_1) - \mathbb{E}(c|\theta_0)
$$

are the *increment in the mean effectiveness* and the *increment in the mean cost*. Obviously, these two functions are the function of $\bf\theta$. Recall that we aim to find $\mathcal{U}^*:= \max\mathcal{U}_t$, i.e. we choose $t=1$ if 

$$
EIB:= \mathcal{U}_1 - \mathcal{U}_0 > 0
(\#eq:3)
$$

thus,

$$
\mathcal{U}^* = \max\{EIB,0\}+\mathcal{U}_0.
$$

## Choosing a utility function: The net benefit

To obtain `r lb(eq.2)` we will need to a pre-defined function of utility $u(y,t)$ through $e$ and $c$, the *net benefit*, for example,

$$
u(y,t) = ke-c,
$$

where $k$ is a willingness-to-pay parameter used to put the cost and benefit on the same scale, and it is the budget the decision-maker is willing to pay to upgrade the benefits by one unit. Let us review the relationship among aforementioned measures: 

:::{.makebox}
- The expected incremental benefit: $EIB:=\mathcal{U}_t-\mathcal{U}_0$;
- The expected utility $\mathcal{U}_t = \mathbb{E}[u(y,t)]$;
- $u(y,t) = ke-c$
- $\Delta_e:= \mathbb{E}(e|\theta_1)-\mathbb{E}(e|\theta_0)$, and $\Delta_c:= \mathbb{E}(c|\theta_1) - \mathbb{E}(c|\theta_0)$
:::

Thus, 

$$
EIB = \mathbb{E}(k\Delta_e-\Delta_c) = k\mathbb{E}(\Delta_e) - \mathbb{E}(\Delta_c),
(\#eq:4)
$$

and because $\Delta_e$ and $\Delta_c$ are expectations of $e|\bf \theta$ and $c|\bf\theta$, by total of expectation, $EIB$ is an expectation on distribution of $\bf\theta$. Let us divide both sides of `r lb(eq.4)` and arrange terms we obtain 

$$
k = \frac{EIB+\mathbb{E}(\Delta_c)}{\mathbb{E}(\Delta_e)} = \frac{EIB}{\mathbb{E}(\Delta_e)}+ICER
$$

thus,

$$
ICER = k-\frac{EIB}{\mathbb{E}(\Delta_e)}
$$

if $EIB >0$ (by `r lb(eq.3)` this is the condition we expect to see), we have

$$
ICER < k,
$$

so interventions having the $ICER$ is less than the willing-to-pay threshold are deemed cost-effective. 

Also, we will need to define an appropriate measures of the cost and effectiveness. The total cost relative to each treatment can be added up the cost of each patient, i.e

$$
c_t := c_t^{drug}(N-SE_t)+ (c_t^{drug}+c^{amb})A_t+(c_t^{drug}+c^{hosp})H_t.
$$

As for the measure of effectiveness, we count the number of patients who do not experience side effects

$$
e_t:= N-SE_t.
$$

# The value of information (VoI)

## Motivation

We shall consider the following table 

```{r tab1}
tab1<-
tibble(decision = c("d1","d2","d3","probability"),
       `state 1` = c(12,15,5,0.5),
       `state 2` = c(9,11,18,0.2), 
       `state 3` = c(13,8,10,0.3)
       )
MyTable(tab1, "current information.")
```

and two measures *expected value with perfect information (EVwPI)* and *expected value without perfect information (EVwoPI)*, which are defined as follows

$$
EVwPI = \max\mathbb{E}(S);
(\#eq:5)
$$
$$
EVwoPI = \mathbb{E}[\max(S)].
(\#eq:6)
$$

Hence, we obtain $\mathbb{E}(S)$ for $d_1$, $d_2$ and $d_3$ are 

```{r}
exp_s<-
sapply(tab1[-4,-1], function(x) as.matrix(tab1[4,-1])%*%x)
exp_s
```

so $\max\mathbb{E}(S)=$ `r max_exp<-max(exp_s);max_exp`. Then we calculate $\mathbb{E}[\max(S)]$

```{r}
exp_max<-
as.matrix(tab1[4,-1])%*%sapply(tab1[-4,-1], max)%>% c()
```

Finally, the so-called the *expected value of perfect information (EVPI)* is the difference between EVwPI and EVwoPI, i.e. $EVwPI - EVwoPI =$ `r exp_max -max_exp`.

The rationale of EVPI is very much similar to these calculations.

## Probabilistic sensitivity analysis (PSA)

We now suppose the prior distribution $p(\bf{\theta}|\mathcal{D})$ is close to one-point distribution at the true value. Hence, the expected utility is 

$$
U(\theta_t) := \mathbb{E}[u(Y,t)|\mathcal{D}] = \int u(y,t)p(y|\theta_t)dy.
$$

Consequently 

























<!-- ------------------------------------------------------------------- -->
(cont.)








